---
title: "Visualisation and Imputation of Missing Values"
author: "Alexander Kowarik (Statistics Austria), Matthias Templ (ZHAW Winterthur)"
date: "July 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(VIM)
data(sleep, package="VIM")
x <- sleep[, -(8:10)]
x[,c(1,2,4,6,7)] <- log10(x[,c(1,2,4,6,7)])
x2 <- sleep[, c("Exp", "Sleep")]
data(tao, package = "VIM")
x3 <- tao[, c("Air.Temp", "Humidity")]
data(chorizonDL, package = "VIM")
x4 <- chorizonDL[,c(15,101:110)]
x5 <- log(chorizonDL[, c(4,5,8,10,11,16:17,19,25,29,37,38,40)])
```

## Outline / R Package
 - Content:
    - Tools for visualization of missing data structures (and imputed values)
    - Tools for imputation
 - Current CRAN version 4.7.0
 - Development version and issue tracking on github [https://github.com/statistikat/VIM](https://github.com/statistikat/VIM)
 - This presentation and the R code [https://github.com/alexkowa/VIM_ISI2017](https://github.com/alexkowa/VIM_ISI2017)
 - [JSS paper on imputation of missing values with VIM, Kowarik, Templ](https://www.jstatsoft.org/article/view/v074i07)
 - [Advances in Data Analysis and Classification paper on visualization with VIM, Templ, Alfons, Filzmoser](https://www.researchgate.net/publication/226283718_Exploring_incomplete_data_using_visualization_techniques)

## Visualisation of Missing Data
- Always important: knowledge about the structure of missing values. Visualisation vs statistical tests.
- literature with focus on visualization of missing data is sparse
- only a few visualization tools missing data
- R package VIM supports the visualization (also with a GUI).

## Aggregation Plots
```{r,echo=TRUE,results='hide'}
aggr(sleep)
```

## Missing Values in Matrix Form
```{r,echo=TRUE,results='hide',warning=FALSE}
matrixplot(x, sortby = "BrainWgt")
```

## Univariate Plots
```{r,echo=TRUE,results='hide',warning=FALSE}
par(mfrow=c(1,2)); histMiss(x2); spineMiss(x3)
```

## Bivariate Plots
```{r,echo=TRUE,results='hide',warning=FALSE}
marginplot(x3)
```

## Multivariate Plots
```{r,echo=TRUE,results='hide',warning=FALSE}
parcoordMiss(x4,plotvars=2:11, interactive = FALSE)
```

## Multiple Plots
```{r,echo=TRUE,results='hide',warning=FALSE}
pbox(x5)
```


## Donor Imputation - hotdeck
```{r,echo=TRUE,eval=FALSE}
hotdeck(data, variable = NULL, ord_var = NULL,
  domain_var = NULL, makeNA = NULL, NAcond = NULL,
  impNA = TRUE, donorcond = NULL, imp_var = TRUE,
  imp_suffix = "imp")
```
 - *data* - data.frame
 - *variable* - variables to be imputed
 - *ord_var* - variables to sort by
 - *domain_var* - variables to build imputation classes
 - a random sort variable is always be added
 
## Donor Imputation - kNN
```{r,echo=TRUE,eval=FALSE}
kNN(data, variable = colnames(data), metric = NULL,
  k = 5, dist_var = colnames(data), weights = NULL,
  numFun = median , catFun = maxCat , makeNA = NULL,
  NAcond = NULL, impNA = TRUE, donorcond = NULL,
  mixed = vector(), mixed.constant = NULL, trace = FALSE ,
  imp_var = TRUE, imp_suffix = "imp", addRandom = FALSE)
```
 - *dist_var* - variables used for distance combination
 - *weights* - weights for distance computation
 - *numFun*, *catFun* - aggregation function for numerical or
categorical target variables (*sampleCat*, *maxCat*).
 - *addRandom* - add a random variable to the distance computation (very low weight)
 
## Donor Imputation - matchImpute
Random within groups imputation, grouping variables are dropped sequentially in case no all values are missing in a group.
```{r,echo=TRUE,eval=FALSE}
matchImpute(data,
  variable = colnames(data)[!colnames(data) %in% match_var],
  match_var, imp_var = TRUE, imp_suffix = "imp")
```
 - *match_var* variables to build groups
 
## Iterative (Robust) Regression Imputation (1)
```{r,echo=TRUE,eval=FALSE}
irmi(x, eps = 5, maxit = 100, mixed = NULL,
  mixed.constant = NULL, count = NULL, step = FALSE ,
  robust = FALSE , takeAll = TRUE, noise = TRUE,
  noise.factor = 1, force = FALSE , robMethod = "MM",
  force.mixed = TRUE, mi = 1, addMixedFactors = FALSE ,
  trace = FALSE , init.method = "kNN")
```
 - *robust* - robust or non-robust
 - *step* - *stepAIC* in every iteration
 - *mixed* - column indices of semi-continuous variables
 - *count* - column indices of count variables (Poisson)
 - *noise* - add a random error to the imputed value
 - *mi* - number of imputations $\Rightarrow$ multiple imputation

## Imputation Benchmarking (1)

```{r,echo=FALSE,results='hide'}
library(ggplot2);library(microbenchmark)
load("timingHD2.RData")
fillup <- function(x,l){
  x <- c(x,rep(NA,l-length(x)))
  x
}
tmed <- do.call("rbind",lapply(mb,function(x)fillup(print(x,unit="s")$median,4)))
tlq <- do.call("rbind",lapply(mb,function(x)fillup(print(x,unit="s")$lq,4)))
tuq <- do.call("rbind",lapply(mb,function(x)fillup(print(x,unit="s")$uq,4)))

out <- data.frame(method=c(rep("matchImpute",7),rep("hotdeck",7),rep("kNN",7),rep("irmi",7)),
                  medianTime=c(tmed[,1],tmed[,2],tmed[,3],tmed[,4]),
                  lq=c(tlq[,1],tlq[,2],tlq[,3],tlq[,4]),
                  uq=c(tuq[,1],tuq[,2],tuq[,3],tlq[,4]),
                  nrow=rep(nRowsAll,4),
                  nmissing=rep(nRowsAll,4)/10)

ggplot(out[out$nrow<=1e+04,],aes(y=medianTime,x=nrow,color=method))+
  geom_ribbon(alpha=.1,color="transparent",aes(fill=method,ymax=uq,ymin=lq))+
  geom_line()
```

## Imputation Benchmarking (2)

```{r,echo=FALSE,results='hide'}
ggplot(out[out$method%in%c("matchImpute","hotdeck"),],aes(y=medianTime,x=nrow,color=method))+
  geom_ribbon(alpha=.1,color="transparent",aes(fill=method,ymax=uq,ymin=lq))+
  geom_line()+
  geom_vline(xintercept = 1e+04,color="grey70")
```

## Iterative Robust Regression Imputation (2)
 ![irmi imputation](cont-semi-impr-cla.pdf "irmi imputation")

